Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants

### 2022-06-16 ms/ms.Rmd
major changes before resubmission

---
output:
  bookdown::pdf_document2:
    includes:
      before_body: ../template/amnat_doc_prefix.tex
      in_header: ../template/amnat_preamble.tex
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
fontsize: 12pt
linestretch: 1.7
link-citations: yes
bst: ../template/amnat.bst
bibliography: stomata-independence.bib
---

@@

```{r}
library(dplyr)
library(glue)
library(lubridate)
library(readr)
library(ropenstomata)
library(stringr)
library(tibble)

@@

@@

# 07_make-pairs ----
pair_div = read_rds("objects/pair_div.rds")
n_pair = nrow(pair_div)

# 10_fit-pairs ----
fit = read_rds("objects/fit.rds")

@@

# 17_summarize-fit ----
mo = read_rds("objects/modeloutput.rds")

@@

<!-- % Covering page info I used for New Phyt -->
<!-- % For Am Nat, moved to "template/amnat_preamble.text" -->
<!-- \begin{flushleft} -->
<!-- {\Large -->
<!-- \textbf\newline{Developmental integration cannot explain major features of stomatal anatomical evolution in seed plants} -->
<!-- } -->
<!-- \newline -->
<!-- \\ -->
<!-- Christopher D. Muir\textsuperscript{1,*}, -->
<!-- Miquel \`{A}ngel Conesa\textsuperscript{2}, -->
<!-- Jeroni Galm\'{e}s\textsuperscript{2}, -->
<!-- Varsha S. Pathare\textsuperscript{3}, -->
<!-- Patricia Rivera\textsuperscript{4}, -->
<!-- Rosana López Rodríguez\textsuperscript{5}, -->
<!-- Teresa Terrazas\textsuperscript{4}, -->
<!-- Dongliang Xiong\textsuperscript{6} -->
<!-- \\ -->
<!-- \bigskip -->
<!-- \bf{1} School of Life Sciences, University of Hawaii at M\=anoa, Honolulu, HI 96822, USA \\ -->
<!-- \bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\ -->
<!-- \bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\ -->
<!-- \bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\ -->
<!-- \bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\ -->
<!-- \bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\ -->
<!-- \bigskip -->
<!-- * cdmuir@hawaii.edu -->

@@

\textit{Manuscript elements}: Figure 1, figure 2, figure 3, figure 4, figure 5, table 1, online appendix A (including Notes A1, table A1, table A2, and table A3).

@@

\textit{Keywords}: adaptation, amphistomy, developmental integration, leaf, phylogenetic comparative methods, stomata

@@

\textit{Word Count}: 4,704

@@

Developmental integration can cause traits to covary over macroevolutionary time and in some cases prevent populations from reaching their adaptive optima. Developmental integration between stomatal size and density may contribute to two major features of stomatal anatomical evolution: inverse size-density scaling and bimodal stomatal ratio. If these patterns result from developmental integration, we predicted that in amphistomatous leaves 1) stomatal size and density should covary similarly on both abaxial and adaxial surfaces and 2) stomatal traits (size and density) on each surface should covary isometrically. We synthesized data on stomatal density and length from amphistomatous leaves of `r n_tax3` terrestrial seed plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from `r n_pair` phylogenetically independent contrasts using a robust Bayesian model. Adaxial stomatal density, but not length, is evolutionarily labile and not strongly integrated with stomatal length or abaxial stomatal density. Hence, developmental integration alone cannot explain inverse size-density scaling nor bimodal stomatal  ratio. Quasi-independent evolution of stomatal anatomical traits facilitates largely unfettered access to fitness optima. If stomatal anatomical traits are near their current fitness optimum, this implies that limits on trait (co)variance result from selective rather than developmental constraints. However, we cannot rule out that developmental integration is important in some lineages. Future research should identify the mechanistic basis of (dis)integration in stomatal development.

@@

The ability for traits to evolve independently of one another is a necessary prerequisite for adaptation to complex environments [@lewontin_adaptation_1978]. If traits can evolve independently and there is sufficient genetic variation, then selection should move populations toward their multivariate phenotypic optimum. Adaptive evolution may be constrained if traits cannot evolve independently because they are developmentally integrated. Developmentally integrated traits have a "disposition for covariation" [@armbruster_integrated_2014], meaning that evolutionary divergence between lineages in one character will be tightly associated with divergence in another character. Allometry is a classic example of developmental integration that may constrain phenotypic evolution (reviewed in @pelabon_evolution_2014). Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection [@lande_quantitative_1979]. However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection [@hansen_is_2003]. For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators (Berg’s rule) [@berg_general_1959; @berg_ecological_1960; @conner_raissa_2014; @armbruster_covariance_1999]. In this study we are interested in quantifying the strength of developmental integration in stomatal anatomy and whether developmental integration might hamper or accelerate adaptive evolution.

Stomata are microscopic pores on the leaf or other photosynthetic surfaces of most land plants formed by a pair of guard cells. Here we limit our focus to stomatal traits on leaves within terrestrial seed plants, primarily angiosperms. The density, size, and patterning of stomata on a leaf set the maximum stomatal conductance to CO$_2$ diffusing into a leaf and the amount of water that transpires from it [@sack_hydrology_2003; @franks_effect_2001; @galmes_leaf_2013]. Plants typically operate below their anatomical maximum by dynamically regulating stomatal aperture. Even though operational stomatal conductance determines the realized photosynthetic rate and water-use efficiency, anatomical parameters are useful in that they set the range of stomatal function [@de_boer_optimal_2016] and are correlated with actual stomatal function under natural conditions [@murray_consistent_2020].

Two salient features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as “inverse size-density scaling” and “bimodal stomatal ratio”  (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata [@weiss_untersuchungen_1865; @franks_maximum_2009; @de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021]. Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. \ref{fig:concepts}b). Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species [@metcalfe_anatomy_1950; @parkhurst_adaptive_1978; @mott_adaptive_1982], especially herbaceous plants from open, high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. @muir_making_2015 described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora (*cf.* @peat_comparative_1994 Fig 1). For both features, we limit our focus in this study to interspecific variation in mean trait values and do not seek to understand intraspecifc variation.

@@

Given the significance of stomata for plant function and global vegetation modeling, we would like to understand whether these major anatomical features are shaped primarily by adaptive or nonadaptive evolutionary forces. We briefly review adaptive hypotheses for both inverse size-density scaling and bimodal stomatal ratio. Then we discuss how developmental integration might contribute to these features, in addition to or in lieu of adaptive evolution. Stomatal size and density determine the maximum stomatal conductance to CO$_2$ and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO$_2$ for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO$_2$. Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference [@zeiger_stomatal_1987], incur metabolic costs [@deans_optimization_2020], and/or risk hydraulic failure [@henry_stomatal_2019]. The distribution of stomata and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. @franks_maximum_2009 further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

Developmental integration between stomatal size and density, mediated by meristematic cell volume, could also explain inverse size-density scaling. If the size of both guard cells and the size of pavement cells between stomata are determined by initial meristematic cell volume, then changes in cell volume early in leaf development would cause both increased stomatal size and lower density [@brodribb_unified_2013]. This type of developmental integration would not hinder adaptive evolution when the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by @franks_maximum_2009, then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

The adaptive significance of variation in stomatal ratio is unknown, but we have some clues based on the distribution of hypo- and amphistomatous leaves. The functional significance of amphistomy is that it adds a second parallel pathway from the substomatal cavities through the leaf internal airspace to sites of carboxylation in the mesophyll [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984]. Thus amphistomatous leaves have lower resistance to diffusion through the airspace which increases the photosynthetic rate [@parkhurst_intercellular_1990]. Despite this amphistomy advantage, most leaves are hypostomatous, suggesting that the benefits of amphistomy in terms of increased photosynthesis usually do not outweigh the costs of stomata on the upper surface. Amphistomy should increase photosynthesis under saturating-light conditions where CO$_2$ supply limits photosynthesis. This may explain why amphistomatous leaves are most common in high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017], especially in herbs [@muir_light_2018]. However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants [@muir_is_2019], suggesting that we need to understand the costs of upper stomata better. Upper stomata increase the susceptibility to rust pathogens in *Populus* [@mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021]. Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits [@buckley_how_2015]. Other hypotheses about the adaptive significance of stomatal ratio are discussed in @muir_making_2015 and @drake_two_2019.

A striking feature of the interspecific variation in stomatal ratio is that trait values are not uniformly distributed, but strongly bimodal. @muir_making_2015 derived general conditions in which bimodality arises because adaptive optima are restricted to separate regimes, but this model has not been tested. An alternative hypothesis is that stomatal traits on the ab- and adaxial surfaces are developmentally integrated because stomatal development is regulated the same way on each surface. In hypostomatous leaves, stomatal development is turned off in the adaxial surface. In amphistomatous leaves, stomatal development proceeds on both surfaces, but evolutionary changes in stomatal development affect traits on both surfaces because they are tethered by a shared developmental program. This model of developmental integration would lead to a bimodal trait distribution because leaves would either be hypostomatous (stomatal ratio equal to 0) or have similar densities on each surface (stomatal ratio approximately 0.5). To our knowledge, this hypothesis has not been put forward in the literature but came about during discussion with one of us (Muir, personal communication with EJ Edwards).

We reasoned that the developmental hypotheses for inverse size-density scaling and bimodal stomatal ratio could be tested using amphistomatous leaves. Since both leaf surfaces are formed from the same genome and meristematic cells, if developmental integration is strong we would expect similar patterns of trait covariation in the abaxial and adaxial surface (see below for specific predictions). Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can be developmentally disintegrated. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection [@barrett_sexual_2013]. Below we reiterate the hypotheses and specific predictions for amphistomatous leaves:

@@

- *Hypothesis*: Meristematic cell volume mediates developmental integration between stomatal size and density

- *Prediction*: Amphistomatous leaves will exhibit identical size-density scaling on each surface

>>>>

---
output:
  bookdown::pdf_document2:
    includes:
      before_body: ../template/amnat_doc_prefix.tex
      in_header: ../template/amnat_preamble.tex
    keep_tex: yes
    latex_engine: pdflatex
    number_sections: no
    toc: no
  bookdown::html_document2:
    number_sections: no
    theme: readable
    toc: yes
  bookdown::tufte_html2:
    number_sections: no
    toc: yes
  bookdown::word_document2: null
fontsize: 12pt
linestretch: 1.7
link-citations: yes
bst: ../template/amnat.bst
bibliography: stomata-independence-1.bib
---

@@

```{r}
library(checkmate)
library(dplyr)
library(glue)
library(lubridate)
library(readr)
library(ropenstomata)
library(stringr)
library(tibble)

@@

# Hack to fix bibliography issues with export from Zotero
read_lines("ms/stomata-independence.bib") |>
  str_replace_all("de Boer", "{de Boer}") |>
  write_lines("ms/stomata-independence-1.bib")

@@

# 07_make-pairs ----
pair_div = read_rds("objects/pair_div.rds")
n_pair = nrow(pair_div)
pair_div_genome = read_rds("objects/pair_div_genome.rds")
n_pair_genome = nrow(pair_div_genome)

# 10_fit-models1 ----
fit12 = read_rds("objects/fit12.rds")

@@

# 17_summarize-fit1 ----
h12o = read_rds("objects/h12output.rds")

# 18_fit-models2 ----
fit3 = read_rds("objects/fit3.rds")

# 19_prepare-plotting2.R ----
h3_hpd = read_rds("objects/h3_hpd.rds")
h3_hpd_density = h3_hpd |>
  filter(name == "diff_density") |>
  mutate(across(where(is.numeric), round, digits = 1)) |>
  transmute(s = glue("{value} [{.lower},{.upper}]"))

h3_hpd_length = h3_hpd |>
  filter(name == "diff_length") |>
  mutate(across(where(is.numeric), round, digits = 1)) |>
  transmute(s = glue("{value} [{.lower},{.upper}]"))

# 21_summarize-fit2 ----
h3o = read_rds("objects/h3output.rds")

# Assert that all details about fit12 and fit3 are identical
assert_true(as.character(fit12$version$cmdstan) == as.character(fit3$version$cmdstan))
assert_true(as.character(fit12$version$brms) == as.character(fit3$version$brms))
assert_true(as.character(fit12$version$cmdstanr) == as.character(fit3$version$cmdstanr))
assert_true(length(fit12$fit@stan_args) == length(fit3$fit@stan_args))
assert_true(fit12$fit@stan_args[[1]][["warmup"]] == fit3$fit@stan_args[[1]][["warmup"]])
assert_true(fit12$fit@stan_args[[1]][["iter"]] == fit3$fit@stan_args[[1]][["iter"]])

@@

<!-- % Covering page info I used for New Phyt -->
<!-- % For Am Nat, moved to "template/amnat_preamble.text" -->
<!-- \begin{flushleft} -->
<!-- {\Large -->
<!-- \textbf\newline{Packing limits and developmental integration are not sufficient to explain stomatal anatomical evolution in flowering plants} -->
<!-- } -->
<!-- \newline -->
<!-- \\ -->
<!-- Christopher D. Muir\textsuperscript{1,*}, -->
<!-- Miquel \`{A}ngel Conesa\textsuperscript{2}, -->
<!-- Jeroni Galm\'{e}s\textsuperscript{2}, -->
<!-- Varsha S. Pathare\textsuperscript{3}, -->
<!-- Patricia Rivera\textsuperscript{4}, -->
<!-- Rosana López Rodríguez\textsuperscript{5}, -->
<!-- Teresa Terrazas\textsuperscript{4}, -->
<!-- Dongliang Xiong\textsuperscript{6} -->
<!-- \\ -->
<!-- \bigskip -->
<!-- \bf{1} School of Life Sciences, University of Hawaiʻi at M\=anoa, Honolulu, HI 96822, USA \\ -->
<!-- \bf{2} Research Group on Plant Biology under Mediterranean Conditions, Departament de Biologia, Universitat de les Illes Balears, Ctra. Valldemossa km 7.5, E-07122, Palma, Spain \\ -->
<!-- \bf{3} School of Biological Sciences, Washington State University, Pullman, WA 99164-4236, USA \\ -->
<!-- \bf{4} Departamento de Botánica, Instituto de Biología, Universidad Nacional Autónoma de México, Apartado Postal 70‑367, 04510 Mexico City, Mexico \\ -->
<!-- \bf{5} Departamento de Sistemas y Recursos Naturales, Universidad Politécnica de Madrid, 28040 Madrid, Spain \\ -->
<!-- \bf{6} National Key Laboratory of Crop Genetic Improvement, MOA Key Laboratory of Crop Ecophysiology and Farming System in the Middle Reaches of the Yangtze River, College of Plant Science and Technology, Huazhong Agricultural University, Wuhan, Hubei 430070, China\\ -->
<!-- \bigskip -->
<!-- * cdmuir@hawaii.edu -->

@@

\textit{Manuscript elements}: Figure 1, figure 2, figure 3, figure 4, figure 5, figure 6, table 1, online appendix A (including Notes A1, notes A2, notes A3, figure A1, figure A2, figure A3, table A1, table A2, and table A3).

@@

\textit{Keywords}: adaptation, amphistomy, developmental integration, leaf, packing limits, phylogenetic comparative methods, stomata

@@

\textit{Word Count}: <!--4,704-->

@@

Functional and developmental constraints on phenotypic variation may cause traits to covary over millions of years and slow populations from reaching their adaptive optima. Alternatively, trait covariation may result from selective constraint if some trait combinations and generally maladaptive. Quantifying the relative contribution of functional, developmental, and selective constraints on phenotypic variation is a longstanding goal of macroevolution, but it is often difficult to distinguish different types of constraints. The stomatal anatomy of leaves with stomata on both surfaces (amphistomatous) present a unique opportunity to test the significance of functional and developmental constraints on phenotypyic evolution at broad phylogenetic scales. The key insight is that stomata on each leaf surface encounter the same functional and developmental constraints, but potentially very different selective constraints. If stomatal traits on each surface evolve differently, this implies that independent evolution is possible and functional or developmental constraints alone likely do not explain covariance between stomatal anatomical traits. Packing limits and cell-size-mediated developmental integration are hypothesized to constrain variation in stomatal anatomy. We tested this synthesizing data on stomatal density and length from amphistomatous leaves of `r n_tax3` terrestrial flowering plant taxa mostly from the literature. We estimated the covariance in divergence between stomatal traits from `r n_pair` phylogenetically independent contrasts using a robust Bayesian model. Contrary to packing limit and developmental integration hypotheses, stomatal anatomy on each surface diverged partially independently. To better understand the global variation in ecologically important traits like stomata will likely require adaptive explanations for what constrains phenotypic (co)variation.

@@

The ability for traits to evolve independently of one another is a necessary prerequisite for adaptation to complex environments [@lewontin_adaptation_1978]. If traits can evolve independently and there is sufficient genetic variation, then selection should move populations toward their multivariate phenotypic optimum. Yet divergence in one trait often covaries with other traits and covariance can persist for millions of years [@schluter_adaptive_1996]. Covariance is "a rough local measure of the strength of constraint" [@maynard_smith_developmental_1985] that can be broken down into functional, developmental, and selective constraints. Functional constraints are "limitations imposed by time, energy, or the laws of physics" [@arnold_constraints_1992]. In other words, certain trait combinations are not physically or geometrically possible. A classic example is shell coiling among invertebrate lineages in which the morphospace of possible phenotypes is constrained by hard geometrical limits [@raup_geometric_1966; @mcghee_theoretical_1999]. Within the space of possible phenotypes, developmental constraints can "bias...the production of variant phenotypes or [place] a limitation on phenotypic variability" [@maynard_smith_developmental_1985]. For example, Fibonacci phyllotaxis may arise from a packing constraint of primordia on the developing apex [@mitchison_phyllotaxis_1977; @reinhardt_law_2022; but see @niklas_role_1988 for an adaptive explanation]. Natural selection constrains phenotypic variation by preventing maladaptive forms from evolving. Selection causes trait covariation because 'missing' trait combinations are maladaptive. Understanding phenotypic constraint is challenging, but a useful starting point is determining whether phenotypic covariation can be explained by functional or developmental constraints [@mcghee_theoretical_1999; @mcghee_geometry_2007; @olson_plant_2019]. If phenotypic covariation is inconsistent with functional and developmental constraints, this provides a strong impetus to test for selective constraint.

In this study, we will address packing problems and developmental integration, specific forms of functional and developmental constraint relevant to our study system, stomatal anatomy. We introduce these concepts generally in this paragraph. Packing a number objects into a finite space is common functional constraint on organisms. Regular geometries that appear in nature such as helices and hexagons (think DNA and honeycombs) are often optimal solutions to packing problems [@mackenzie_proving_1999; @maritan_optimal_2000]. Notice that functional constraint does not preclude selection, but the presence of a packing limit changes the range of possible phenotypes. Developmental integration is a form of developmental constraint on multivariate phenotypic evolution and we use them interchangeably in this study. Developmentally integrated traits have a "disposition for covariation" [@armbruster_integrated_2014], meaning that evolutionary divergence between lineages in one trait will be tightly associated with divergence in another trait. Allometry is a classic, albeit contested, example of developmental integration that may constrain phenotypic evolution [reviewed in @pelabon_evolution_2014]. Strong allometric covariation between traits within populations can constrain macroevolutionary divergence for long periods of time depending on the strength and direction of selection [@lande_quantitative_1979]. However, developmental integration does not necessarily hamper adaptation, and can even accelerate adaptive evolution when trait covariation is aligned with the direction of selection [@hansen_is_2003]. For example, fusion of floral parts increases their developmental integration which may increase the rate and precision of multivariate adaptation to specialist pollinators [Berg’s rule, @berg_general_1959; @berg_ecological_1960; @conner_raissa_2014; @armbruster_covariance_1999].

Biologists have studied phenotypic constraints for decades, but progress is challenging because many multivariate phenotypes are too complex or poorly understood to quantitatively distinguish functional, developmental, and selective constraints over macroevolutionary timescales. Stomatal anatomy on the leaves of land plants provide an exceptional opportunity because 1) there are 1000s of species to compare and 2) the main packing constraints and developmental steps are analytically tractable. This means it is possible to derive quantitative predictions and test their generality using large comparative data sets representing millions of year of evolutionary history. A heretofore unappreciated fact about stomata is that many leaves have stomata on both lower and upper surfaces. The packing and developmental constraints are the same for stomata on each surface, but the selective constraints may potentially differ. Therefore, if packing and developmental constraints dominate, stomatal anatomy on each surface should diverge in concert. Failure to do so implies that independent evolution is possible and that selective constraints most likely explain at least some of the covariance between traits. Phylogenetic comparisons of stomatal anatomy provide a statistically powerful, general, and elegant way to distinguish different phenotypic constraints that would be impossible in many other traits with as much ecological importance. The next sections provides background information on stomatal anatomy, how it varies, and why functional or developmental constraints might be important. Readers not interested in this background may skip ahead to hypotheses and predictions.

## The adaptive significance and ecological distribution of variation in stomatal anatomy

Stomata are microscopic pores formed by a pair of guard cells that regulate gas exchange (CO$_2$ gain and water vapor loss) on the leaf or other photosynthetic surfaces of most land plants. Stomata originated once in the history of land plants around 500 Ma, diversified rapidly in density and size, and have been maintained in most lineages except some bryophytes and aquatic plants [recently reviewed @clark_origin_2022]. Stomata respond physiologically by opening and closing in response to light, humidity, temperature, circadian rhythm, and plant water status [@hetherington_role_2003; @lawson_guard_2020]. On this short time scale, the stomatal size, density, and distribution on a mature leaf do not change, so the maximum rate of is fixed. At somewhat longer timescales, the plant may respond to environmental cues such as light and CO$_2$ by altering stomatal anatomy in new leaves [@casson_influence_2008]. Physiological responses (aperture change) and plastic responses (new leaves with changed anatomy) may be alternative strategies for plants to acclimate to environmental change [@haworth_co-ordination_2013]. Finally, stomatal anatomy can evolve due to inherited changes in stomatal development. Plastic and genetic changes in stomatal anatomy are likely both ecologically important, but most studies do not use a common garden design that would tease apart their relative contribution.

We focus on anatomical variation in the density, size, and patterning of stomata on a leaf because these factors set the maximum stomatal conductance to CO$_2$ diffusing into a leaf and the amount of water that transpires from it [@sack_hydrology_2003; @franks_effect_2001; @galmes_leaf_2013; @harrison_influence_2020]. Plants typically operate below their anatomical maximum by dynamically regulating stomatal aperture. Even though operational stomatal conductance determines the realized photosynthetic rate and water-use efficiency, anatomical parameters are useful in that they set the range of stomatal function [@de_boer_optimal_2016] and are correlated with actual stomatal function under natural conditions [@murray_consistent_2020]. All else being equal, larger, more densely packed, but evenly spaced stomata increase gas exchange [@franks_maximum_2009; @dow_physiological_2014; @lehmann_effects_2015]. Smaller stomata may also be able to respond more rapidly than larger stomata, proving the ability of leaves to track short duration environmental change [@drake_smaller_2013]. Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in some species [@metcalfe_anatomy_1950; @parkhurst_adaptive_1978; @mott_adaptive_1982]. Amphistomatous leaves have a second parallel pathway from the substomatal cavities through the leaf internal airspace to sites of carboxylation in the mesophyll [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984]. Thus amphistomatous leaves have lower resistance to diffusion through the airspace which increases the photosynthetic rate [@parkhurst_intercellular_1990]. If total stomatal and other conductances to CO$_2$ supply could be held constant, then an amphistomatous leaf will have a greater conductance than am otherwise identical hypostomatous leaf. The magnitude of the advantage depends on upon the resistance to diffusion through the internal airspace, which is variable among species.

The adaptive significance and ecological distribution of leaves with different stomatal anatomies is complex and there is much yet to learn. Seed plants posses a wider range of stomatal anatomies than ferns and fern allies, which are restricted to having large stomata, at low density, only on the lower surface [@de_boer_optimal_2016]. In general, trees and shrubs have greater stomatal density than herbs, but there is a of lot variation within growth forms depending on the ecological niche [@salisbury_i_1928; @kelly_plant_1995]. A commonly observed trend is that leaves from higher light environments tend to have greater stomatal density [@salisbury_i_1928; @gibson_structure-function_1996]. This may explain why, perhaps counterintuitively, plants in dry environments tend to have more stomata. Drier habitats are more open, enabling plants with higher stomatal density can photosynthesize more when water is available, but close stomata during drought [@liu_variation_2018]. Over recent human history, stomatal density has tended to decline within species as atmospheric CO$_2$ concentrations have risen [@woodward_stomatal_1987; @royer_stomatal_2001]. It is unclear whether most of this change is plastic or genetic, but the overall direction is consistent with the hypothesis that plants decrease gas exchange as CO$_2$ availability increases.

The adaptive significance of variation in stomatal ratio is uncertain, but we have some clues based on the distribution of hypo- and amphistomatous leaves. Despite the fact that amphistomy can increase photosynthesis, most leaves are hypostomatous, suggesting that the benefits of amphistomy. Amphistomy should increase photosynthesis most under saturating-light conditions where CO$_2$ supply limits photosynthesis. This may explain why amphistomatous leaves are most common in high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017], especially in herbs [@muir_light_2018]. However, the light environment alone cannot explain why hypostomatous leaves predominate in shade plants [@muir_is_2019], suggesting that we need to understand the costs of upper stomata better. Upper stomata increase the susceptibility to rust pathogens in *Populus* [@mckown_association_2014; @mckown_role_2019; @fetter_growthdefense_2021]. Amphistomy may also cause the palisade mesophyll to dry out under strong vapor pressure deficits [@buckley_how_2015]. Other hypotheses about the adaptive significance of stomatal ratio are discussed in @muir_making_2015 and @drake_two_2019.

## Major features of stomatal anatomical macroevolution

Two salient features of stomatal anatomy have been recognized for decades but we do not yet understand the evolutionary forces that generate and maintain them. We denote these two features as “inverse size-density scaling” and “bimodal stomatal ratio” (Fig. \ref{fig:concepts}). Inverse size-density scaling refers to the negative interspecific correlation between the size of the stomatal apparatus and the density of stomata [@weiss_untersuchungen_1865; @franks_maximum_2009; @de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021]. Across species, leaves with smaller stomata tend to pack them more densely, but there is significant variation about this general trend (Fig. \ref{fig:concepts}a). Stomatal size and density determine the maximum stomatal conductance to CO$_2$ and water vapor but also take up space on the epidermis, which could be costly for both construction and maintenance. Natural selection should favor leaves that have enough stomata of sufficient size to supply CO$_2$ for photosynthesis. Hence leaves with few, small stomata and high photosynthetic rates do not exist because they would not supply enough CO$_2$. Conversely, excess stomata or extra large stomata beyond the optimum may result in stomatal interference where the CO$_2$ concentration gradient around one stomate merges with that of its neighbor [@zeiger_stomatal_1987; @lehmann_effects_2015], incur metabolic costs [@deans_optimization_2020], and/or risk hydraulic failure [@henry_stomatal_2019]. The distribution of stomata and density may therefore represent the combinations that ensure enough, but not too much, stomatal conductance. @franks_maximum_2009 further hypothesized that the evolution of small stomata in angiosperms enabled increased stomatal conductance while minimizing the epidermal area allocated to stomata.

A striking feature of the interspecific variation in stomatal ratio is that trait values are not uniformly distributed, but strongly bimodal \ref{fig:concepts}b. Bimodal stomatal ratio refers to the observation that the ratio of stomatal density on the adaxial (upper) surface to the density on the abaxial (lower) has distinct modes (Fig. \ref{fig:concepts}b). Amphistomy occurs most often in herbaceous plants from open, high light habitats [@salisbury_i_1928; @mott_adaptive_1982; @gibson_structure-function_1996; @smith_associations_1998; @jordan_using_2014; @muir_making_2015; @bucher_stomatal_2017; @muir_light_2018]. @muir_making_2015 described bimodal stomatal ratio formally but the pattern is apparent in earlier comparative studies of the British flora (*cf.* @peat_comparative_1994 Fig 1).

@@

## Packing limits, developmental integration, and stomatal anatomy

Given the significance of stomata for plant function and global vegetation modeling [@berry_stomata:_2010], we would like to understand what factors constrain their anatomical variation. Could packing limits and developmental integration explain? We limit our focus in this study to interspecific variation in mean trait values and do not study understand intraspecifc variation here.

Packing constraints and developmental integration could explain inverse size-density scaling. The density and size of a stomata on a planar leaf surface can be viewed as a packing problem where the total area allocated to stomata cannot exceed the total leaf area. This is a functional, or geometric, constraint because certain combinations of large size and high density are not physically possible. This functional constraint cannot explain why combinations of low density and small size are rare, but may explain why stomatal size must decrease when density increase as the leaf runs out of space. The packing limit of functional stomata is less than the entire leaf area, but the exact value is unclear. The realized upper limit is close to 1/3 or 1/2 [@de_boer_optimal_2016; @sack_developmental_2016; @liu_scaling_2021].

Guard cell size and spacing between stomata (the inverse of density) are developmentally intertwined because guard cells and epidermal pavement cells between stomata develop from the same meristem. Before guard cell meristemoids form via asymmetric cell division [@dow_patterning_2014], the size of guard and epidermal cells are influenced by meristematic cell volume and expansion. Evolutionary shifts in meristematic cell volume or expansion rate could cause both increased stomatal size and lower density because epidermal cells between stomata are larger [@brodribb_unified_2013]. For example, larger genomes increase meristematic cell volume [@simova_geometrical_2012], setting a lower bound on final cell volume. Although different expansion rates in guard and epidermal pavement cells can reduce the correlation in their final size, the fact that species with larger genomes tend toward having larger stomata and lower density may indicate an effect of development integration of stomatal anatomy [@beaulieu_genome_2008; @simonin_genome_2018; @roddy_scaling_2020]. Developmental integration in this case would not hinder adaptive evolution when the main axes of selection were aligned with the developmental correlation. For example, if higher maximum stomatal conductance were achieved primarily by increasing stomatal density and decreasing stomatal size as proposed by @franks_maximum_2009, then developmental integration might accelerate the response to selection compared to a case where stomatal size and density are completely independent.

@muir_making_2015 derived general conditions in which bimodality arises because adaptive optima are restricted to separate regimes, but this model has not been tested. An alternative hypothesis is that stomatal traits on the ab- and adaxial surfaces are developmentally integrated because stomatal development is regulated the same way on each surface. In hypostomatous leaves, stomatal development is turned off in the adaxial surface. In amphistomatous leaves, stomatal development proceeds on both surfaces, but evolutionary changes in stomatal development affect traits on both surfaces because they are tethered by a shared developmental program. This is a developmental constraint because the fact that stomatal development is the same on each surface constrains the type of variation available for selection. Developmental integration would lead to a bimodal trait distribution because leaves would either be hypostomatous (stomatal ratio equal to 0) or have similar densities on each surface (stomatal ratio approximately 0.5). To our knowledge, this hypothesis has not been put forward in the literature.

## Hypotheses and predictions

The overarching question is whether major features of stomatal anatomy in terrestrial angiosperms are consistent with packing constraints and developmental integration mediated by cell size. Since stomata on both surfaces of amphistomatous leaves are subject to the same functional and developmental constraints, if these constraints are most important we predict similar patterns of trait covariation on abaxial and adaxial surfaces. Conversely, if traits covary differently on each surface it would indicate that stomatal anatomical traits can evolve independently and selective constraints like contribute to covariation. Analogously, variation in the genetic correlation and interspecific divergence of sexually dimorphic traits in dioecious species demonstrate that integration is not fixed and can be modified by selection [@barrett_sexual_2013]. We framed specific hypotheses and predictions around how functional or developmental constraints might explain either inverse size-density scaling or bimodal stomatal ratio.

@@

Both packing limits and developmental integration could contribute to inverse size-density scaling. If limits on the fraction of epidermal area occupied by stomata constrains that combinations of stomatal size and density that are evolutionarily accessible, then we predict that evolutionary divergence in stomatal size and/or density will decrease as the fraction of epidermal area occupied by stomata increases. Furthermore, if divergence slows as epidermal area occupied by stomata because of a packing limit, it should slow down the same way for both ab- and adaxial surfaces.
